---
title: "Comparison of E(z) mutant brain expression profiles and age-correlated probesets indicates a \"younger\" brain"
author: "Jason Kennerdell"
date: "9/25/2017"
output: pdf_document
---
```{r, "needed packages", include=FALSE, warning=FALSE, message=FALSE}
library(DESeq2); library(ggplot2); library(grid); library(gridExtra); library(genefilter)
```

### Input the data and metadata:
```{r, "input sampleTable", warning=FALSE, message=FALSE, cache=T}
inpath <- "~/Desktop/brain"
outpath <- "~/Desktop/brain/_8C_orderedList"
setwd(inpath)
files <- list.files()
htseq_files <- files[grepl("^JKL.*txt$", files)]
sampleNames <- read.csv("EzSampleNames.csv")
sampleTable <- data.frame(fileName = htseq_files,
                          stringsAsFactors=FALSE)
sampleTable$Library <- gsub("-counts.txt", "", sampleTable$fileName)
sampleTable$Library <- gsub("b", "", sampleTable$Library)
sampleTable$seq.batch <- ifelse(grepl("b", sampleTable$fileName), "B", "A")
sampleTable$seq.batch <- paste(sampleTable$Library, sampleTable$seq.batch)
sampleTable <- merge(sampleTable[,c(1,3)], sampleNames, by = "seq.batch")
sampleTable$genotype <- gsub("-.*$", "", sampleTable$Sample)
sampleTable$genotype <- factor(sampleTable$genotype, levels = c("w", "Ez"))
sampleTable$Temp <- gsub("[A-Z, a-z, -]", "", sampleTable$Sample)
sampleTable$Temp <- factor(sampleTable$Temp, levels = c("25", "29"))
sampleTable$rep <- gsub("^.*-", "", sampleTable$Sample)
sampleTable$condition <- paste(sampleTable$genotype, sampleTable$Temp, sep = "-")
sampleTable
```

### Set up the statistical model
```{r, "design"}
design <- formula(~ Temp + genotype)
```

## DESEq2 Statistics
```{r, "statistics", cache=TRUE, message=FALSE}
dds <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable, directory = inpath, design = design)
# Combine the technical replicates (different runs) by adding the count
# totals for each gene across the two runs:
dds <- collapseReplicates(dds, groupby=dds$Library, run = dds$batch)
dds <- DESeq(dds, betaPrior=T)
# What does the data look like?
head(assay(dds)) # This is the sum of the two runs HTSeq-count output!
# What are the columns?
colData(dds)
results <- results(dds, alpha=0.05)
results$ensembl <- rownames(results)
```

### Prepare MA plots:
```{r, cache=T}
# For Maximum likelihood estimates:
resultsMLE <- results(dds, addMLE=TRUE, alpha = 0.05)
par(mfrow=c(1,2))
plotMA(resultsMLE, MLE=TRUE, alpha = 0.05, main="unshrunken LFC", ylim=c(-10,10))
plotMA(results, alpha = 0.05, main="shrunken LFC", ylim=c(-5,8))
```

### Add Annotation
```{r, cache=T}
# Add usefull gene names:
library(biomaRt)

# Add FBgn Names
mart = useMart("ENSEMBL_MART_ENSEMBL", host="aug2017.archive.ensembl.org")
#listDatasets(mart)
mart = useMart("ENSEMBL_MART_ENSEMBL", host="aug2017.archive.ensembl.org",
               dataset = "dmelanogaster_gene_ensembl")
genemap <- getBM(attributes = c("ensembl_gene_id", "entrezgene", "external_gene_name", "flybasecgid_gene"), filters = "ensembl_gene_id", values = results$ensembl, mart = mart)
idx <- match(results$ensembl, genemap$ensembl_gene_id)
results$entrez <- genemap$entrezgene[idx]
results$geneSymbol <- genemap$external_gene_name[idx]
results$cg <- genemap$flybasecgid_gene[idx]
```

## Prepare Ranked lists of genes for OrderedList analysis
```{r}
# Using age correlated genes by p value and beta coefficient:
correl_lst <- read.csv("~/Desktop/brain/GSE25007/ranked_aging_correlated_genes_pvalue_only.csv")
# How many genes are there in this list?
dim(correl_lst)[1]
results_lst <- as.data.frame(results)
results_lst <- results_lst[!is.na(results_lst$pvalue),]
results_lst$product <- -log10(results_lst$pvalue)*results_lst$log2FoldChange
results_lst <- results_lst[order(results_lst$product),]
# How many genes are there in this list?
dim(results_lst)[1]

# Select only members of lists that are present in the opposite list
results_lst <- results_lst[results_lst$ensembl %in% correl_lst$ensembl,]
head(results_lst)
tail(results_lst)
# How many genes are there in this list?
dim(results_lst)[1]

correl_lst <- correl_lst[correl_lst$ensembl %in% results_lst$ensembl,]
head(correl_lst)
tail(correl_lst)
# How many genes are there in this list?
dim(correl_lst)[1]
```

# OrderedList analysis
## Test for overlap of Positively age correlated genes and down regulated in E(z) genes, one sided:
```{r}
library(OrderedList)
# Test for overlap of age correlated genes and down regulated in E(z) genes, one sided:
myComp <- compareLists(ID.List1=results_lst$ensembl, 
                       ID.List2=correl_lst$ensembl, 
                       mapping=NULL, two.sided = F)
myComp
myOverlap <- getOverlap(myComp, max.rank=500, percent=.95)
#par(mfrow=c(1,2))
plot(myOverlap, no.title=TRUE, no.legend=TRUE, 
     main = "Positively Age Correlated genes vs. \n E(z) downregulated genes")
legend(0,67,legend=c("Overlap", "Shuffle 95% confidence interval"), 
       col=c("black", "orange"), lty=c(1,1))
plot(myOverlap, "scores")
myOverlap

# Make a publication plot:
#########################################################################
jpeg(file=paste(outpath, "Fig_8c.jpg", sep="/"), 
     quality=100,
     res=300,
     width=1440, 
     height=960)
par(cex.axis=1.25, cex.lab=2, mar=c(4,5,4,2), col.lab="white")
plot(myOverlap, no.title=TRUE, no.legend=TRUE, cex.lab=2, 
     #xlab = NA,#Overlapping gene count",
     #ylab = NA,#Rank of genes in the two lists",
     main = "", las=1)
legend(0,67,legend=c("Overlap", "Shuffle 95% confidence interval"), 
       col=c("black", "orange"), lty=c(1,1), cex=.75)
dev.off()
##########################################################################
```

## Test for overlap of age correlated genes and down regulated in E(z) genes, two sided:
```{r}
myComp2 <- compareLists(ID.List1=results_lst$ensembl, 
                        ID.List2=correl_lst$ensembl, 
                        mapping=NULL, two.sided = T)
myComp2
myOverlap2 <- getOverlap(myComp2, max.rank=500, percent=.95)
par(mfrow=c(1,2))
plot(myOverlap2, no.title=TRUE, no.legend=TRUE, 
     main = "Age Correlated genes vs. \n E(z) genes")
legend(0,67,legend=c("Overlap", "Shuffle 95% confidence interval"), 
       col=c("black", "orange"), lty=c(1,1))
plot(myOverlap2, "scores")
myOverlap2
```

## Test for overlap of Negatively age correlated genes and up regulated in E(z) genes, one sided:
```{r}
# Test for overlap of age correlated genes and down regulated in E(z) genes, one sided:
myComp3 <- compareLists(ID.List1=rev(results_lst$ensembl), 
                       ID.List2=rev(correl_lst$ensembl), 
                       mapping=NULL, two.sided = F)
myComp3
myOverlap3 <- getOverlap(myComp3, max.rank=500, percent=.95)
par(mfrow=c(1,2))
plot(myOverlap3, no.title=TRUE, no.legend=TRUE, 
     main = "Negatively Age Correlated genes vs. \n E(z) upregulated genes")
legend(0,52,legend=c("Overlap", "Shuffle 95% confidence interval"), 
       col=c("black", "orange"), lty=c(1,1))
plot(myOverlap3, "scores")
myOverlap3

# Make a publication plot:
jpeg(file=paste(outpath, "Fig_S5f.jpg", sep="/"), width=360, height=240)
par(cex.axis=1.25, cex.lab=2, mar=c(4,5,4,2), col.lab="white")
plot(myOverlap3, no.title=TRUE, no.legend=TRUE, cex.lab=2,
     main = "", las=1)
legend(0,52,legend=c("Overlap", "Shuffle 95% confidence interval"), 
       col=c("black", "orange"), lty=c(1,1), cex=.75)
dev.off()
```

## Test for overlap of age correlated genes and up regulated in E(z) genes, two sided:
```{r}
# Test for overlap of age correlated genes and down regulated in E(z) genes, one sided:
myComp4 <- compareLists(ID.List1=rev(results_lst$ensembl), 
                       ID.List2=rev(correl_lst$ensembl), 
                       mapping=NULL, two.sided = T)
myComp4
myOverlap4 <- getOverlap(myComp4, max.rank=500, percent=.95)
par(mfrow=c(1,2))
plot(myOverlap4, no.title=TRUE, no.legend=TRUE, 
     main = "Age Correlated genes vs. \n E(z) genes")
legend(0,67,legend=c("Overlap", "Shuffle 95% confidence interval"), 
       col=c("black", "orange"), lty=c(1,1))
plot(myOverlap4, "scores")
myOverlap4
```

### Prepare a ranked list for publication: (Supplemental Data 2)
```{r}
ranked_list <- correl_lst[,c(2,4)]
rownames(ranked_list) <- 1:dim(ranked_list)[1]
write.csv(file=file.path(outpath, "rankedAgingGenes.csv"), ranked_list)
```

Save image
```{r}
save.image(file=file.path(outpath, "OrderedList.RData"))
```




